Channel Shaping Using Beyond Diagonal Reconfigurable Intelligent Surface:\\Analysis, Optimization, and Enhanced Flexibility

### 2024-07-19 main.tex
:pencil: Update manuscript

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{d}{D}{Diagonal}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}
\newacronym{los}{LoS}{Line-of-Sight}
\newacronym{csi}{CSI}{Channel State Information}
\newacronym{cscg}{CSCG}{Circularly Symmetric Complex Gaussian}
\newacronym{sca}{SCA}{Successive Convex Approximation}
\newacronym{kkt}{KKT}{Karush-Kuhn-Tucker}

\begin{document}
\title{Channel Shaping Using Beyond-Diagonal Reconfigurable Intelligent Surface:\\Analysis, Optimization, and Implications}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

@@

	and adopt a \gls{bd} architecture featuring element-wise connections for signal amplitude and phase control.

@@

	Of particular interest, \gls{bd}-\gls{ris} is shown to activate multi-stream transmission at lower transmit power than \gls{d}-\gls{ris}, hence attaining the full \gls{dof} at lower \gls{snr} thanks to its higher flexibility of shaping the distribution of channel singular values.

@@

		% pioneer
		Third, we provide a numerical answer to the shaping question by characterizing the Pareto frontiers of channel singular values.
		The enclosed region generalizes most relevant metrics and provides an intuitive singular value shaping benchmark.
		Results show that increasing \gls{bd}-\gls{ris} group size enlarges this region, improving the dynamic range of all singular values and the trade-off in manipulating them.
		% enlarges the singular value region, improving both dynamic range of and trade-off between singular values.

@@

		Fourth, we provide an analytical answer to the shaping question in two special channel conditions.
		For rank-deficient forward/backward channel, we derive asymptotic singular value bounds applying to \gls{d}- and \gls{bd}-\gls{ris}.
		For blocked direct channel, we derive singular value bounds applying to fully-connected \gls{bd}-\gls{ris} with arbitrary number of elements.
		Those bounds are validated by comparing with numerical results above.
		Results show that for a fixed number of elements, \gls{bd}-\gls{ris} can approach the asymptotic bounds better than \gls{d}-\gls{ris}.

@@

		Fifth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem through a local-optimal \gls{ao} and a low-complexity shaping-inspired approach.
		The former updates active beamforming by eigenmode transmission and passive beamforming by geodesic \gls{rcg} until convergence.
		The latter shapes the channel for maximum power gain and then performs eigenmode transmission.
		Interestingly, the rate gap in between diminishes as the \gls{ris} evolves from diagonal to fully-connected.
		We highlight that the power and rate gains of \gls{bd}-\gls{ris} over \gls{d}-\gls{ris} increase with the number of scattering elements and \gls{mimo} dimensions.
		Besides, \gls{bd}-\gls{ris} can help to activate more streams and better approach the \gls{dof} than \gls{d}-\gls{ris} at low transmit \gls{snr}.
		% \gls{bd}-\gls{ris} is shown to activate multi-stream transmission at lower transmit power than \gls{d}-\gls{ris}, hence attaining the full \gls{dof} at lower \gls{snr} thanks to its higher flexibility of shaping the distribution of channel singular values.

@@

		When it comes to \gls{mimo}, each \gls{d}-\gls{ris} element can only apply a common phase shift to the rank-1 indirect channel matrix passing through itself.

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{d}{D}{Diagonal}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}
\newacronym{los}{LoS}{Line-of-Sight}
\newacronym{csi}{CSI}{Channel State Information}
\newacronym{cscg}{CSCG}{Circularly Symmetric Complex Gaussian}
\newacronym{sca}{SCA}{Successive Convex Approximation}
\newacronym{kkt}{KKT}{Karush-Kuhn-Tucker}

\begin{document}
\title{Channel Shaping Using Beyond Diagonal Reconfigurable Intelligent Surface:\\Analysis, Optimization, and Enhanced Flexibility}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

@@

	and adopt a \gls{bd} architecture not restricted to diagonal scattering matrix and featuring element-wise connections for extra wave manipulation flexibility.

@@

	Of particular interest, \gls{bd}-\gls{ris} is shown to activate multi-stream transmission at lower transmit power than \gls{d}-\gls{ris}, hence better approaching \gls{dof} at low \gls{snr} thanks to its higher flexibility of shaping the distribution of channel singular values.

@@

		% pioneer
		Third, we provide a numerical answer to the shaping question by characterizing the Pareto frontiers of channel singular values.
		The enclosed region generalizes most relevant metrics and provides an intuitive shaping benchmark.
		Results show that increasing \gls{bd}-\gls{ris} group size enlarges this region, improving the dynamic range of all singular values and the trade-off in manipulating them.
		% enlarges the singular value region, improving both dynamic range of and trade-off between singular values.

@@

		Fourth, we provide an analytical answer to the shaping question in two special channel conditions.
		For rank-deficient forward/backward channel, we derive singular value bounds applying to \gls{d}- and \gls{bd}-\gls{ris} with asymptotically large number of elements.
		For blocked direct channel, we derive singular value bounds applying to fully-connected \gls{bd}-\gls{ris} with arbitrary number of elements.
		Those bounds are validated by comparing with the numerical results above.
		Results show that for a fixed number of elements, \gls{bd}-\gls{ris} can approach the asymptotic bounds better than \gls{d}-\gls{ris}.

@@

		Fifth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} and a low-complexity shaping-inspired approach.
		The former iteratively updates active beamforming by eigenmode transmission and passive beamforming by geodesic \gls{rcg}, until convergence.
		The latter exploits the \gls{bd}-\gls{ris} to shape the channel for maximum power gain then performs eigenmode transmission.
		% simply shapes the channel by \gls{bd}-\gls{ris} for maximum power gain and performs eigenmode transmission.
		Interestingly, the rate deficit from the shaping-inspired approach diminishes as the \gls{bd}-\gls{ris} evolves towards fully-connected.
		% gap in between diminishes as the \gls{ris} evolves from diagonal to fully-connected.
		We conclude that: 1) channel shaping decouples joint beamforming for reduced design complexity; 2) the power and rate gains of \gls{bd}-\gls{ris} over \gls{d}-\gls{ris} increase with the number of scattering elements and \gls{mimo} dimensions; 3) at low transmit power, \gls{bd}-\gls{ris} can activate more streams and better approach \gls{dof} than \gls{d}-\gls{ris}, thanks to its higher flexibility of shaping the distribution of channel singular values.
		% Interestingly, the rate gap in between diminishes as the \gls{ris} evolves from diagonal to fully-connected.
		% Channel shaping helps to decouple joint beamforming by first designing the channel and then the transmitter.

		% We highlight that the power and rate gains of \gls{bd}-\gls{ris} over \gls{d}-\gls{ris} increase with the number of scattering elements and \gls{mimo} dimensions.
		% Besides, \gls{bd}-\gls{ris} can help to activate more streams and achieves higher \gls{dof} than \gls{d}-\gls{ris} at low transmit \gls{snr}.
		% \gls{bd}-\gls{ris} is shown to activate multi-stream transmission at lower transmit power than \gls{d}-\gls{ris}, hence attaining the full \gls{dof} at lower \gls{snr} thanks to its higher flexibility of shaping the distribution of channel singular values.

@@

		When it comes to \gls{mimo}, each \gls{d}-\gls{ris} element can only apply a scalar phase shift to the associated indirect channel matrix of size $N_\mathrm{R} \times N_\mathrm{T}$ defined in \eqref{eq:channel_equivalent}.
